---
title: "AARAN, Somalia"
subtitle: "Land Restoration Prioritization Report"
author: "Ivy Bartonjo and Tor-G. Vågen"
date: "September 2, 2026"
toc: true
format:
  typst:
    template-partials:
      - typst-template.typ
      - typst-show.typ
    fig-format: png
  html:
    self-contained: true
    code-tools: false
execute:
  echo: false
  warning: false
  message: false
---

```{r}
library(terra)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

output_dir <- "outputs_report"
figure_dir <- "report_pngs"

effort_file <- file.path(output_dir, "AARAN_restoration_effort.tif")
cluster_file <- file.path(output_dir, "AARAN_ecosystem_GMM_clusters.tif")
prob_files <- c(
  Low = file.path(output_dir, "AARAN_effort_prob_low.tif"),
  Medium = file.path(output_dir, "AARAN_effort_prob_medium.tif"),
  High = file.path(output_dir, "AARAN_effort_prob_high.tif")
)
importance_pngs <- c(
  Low = file.path(figure_dir, "AARAN_importance_low.png"),
  Medium = file.path(figure_dir, "AARAN_importance_medium.png"),
  High = file.path(figure_dir, "AARAN_importance_high.png")
)

effort_labels <- c("Low restoration effort", "Medium restoration effort", "High restoration effort")
effort_cols <- c("#2A9D8F", "#E9C46A", "#E76F51")
mclust_reclass <- data.frame(cluster = 1:3, effort_class = c(2, 1, 3))

effort_raster <- function() {
  if (file.exists(cluster_file)) {
    out <- terra::classify(rast(cluster_file), as.matrix(mclust_reclass))
    levels(out) <- data.frame(ID = 1:3, restoration_effort = effort_labels)
    return(out)
  }
  if (file.exists(effort_file)) return(rast(effort_file))
  NULL
}
```

{{< pagebreak >}}

## Introduction

This report presents the results of land health analysis and assessment of restoration effort to support the prioritisation of land restoration interventions in Somalia for the AARAN project. Restoration effort is assessed using soil and land health indicator maps, which are clustered into low, medium, and high restoration effort, respectively. The classes are derived from a model-based clustering workflow using five biophysical indicators: tree cover, erosion, root depth restriction at 50 cm, soil pH, and soil organic carbon (SOC).

The purpose of the report is to give an overview at country, watershed and cluster level, and to provide a summary of the indicator values associated with each restoration effort class. The report is intended for use by AARAN and other stakeholders to inform restoration planning and prioritization.

::: {.callout-note icon=false}
## Key Findings Summary
- **National Effort Distribution**: Approximately **41%** Medium restoration effort, **38%** Low restoration effort, and **21%** High restoration effort.
- **Biophysical Clustering**: Five indicators (tree cover, erosion, RDR50, soil pH, and SOC) grouped into 3 distinct operational management classes via Gaussian mixture modeling (`mclust`).
- **Primary Predictive Drivers**: Tree cover and soil pH consistently show the highest predictive influence across class-specific Random Forest probability models.
:::

## Methods

Randomly generated points were generated and raster values for tree cover, erosion, RDR50, pH, and SOC were extracted for each point. Sampled indicator values were standardized and clustered into three groups using R Statistics through the `mclust` library. This approach groups multi-dimensional biophysical indicators into distinct classes based on underlying probabilistic Gaussian mixture models.

The three resulting classes were manually interpreted as restoration effort classes based on the distribution of indicator values in each cluster. The cluster (group) with high erosion, low SOC, and acidic or alkaline pH was classified as high effort, while the cluster with low erosion, higher SOC and neutral pH was classified as low effort. The cluster with intermediate indicator values was classified as moderate effort. Separate random forest models were then used to represent each class against the same indicator set, producing class specific probability surfaces and variable importance graphs.

The reporting workflow uses three output groups:

- Mclust cluster and restoration effort rasters
- Random forest probability rasters for low, medium, and high effort
- Variable-importance graphs for low, medium, and high effort surfaces

## Results

The Mclust model produced three restoration-effort groups. **Add figure and discuss this, then move on to the table and the cluster assignments.**

These groups provide the main classification used in the report. @tbl-mclust-restoration should be interpreted alongside the map and RF outputs rather than as a final ecological ranking on its own.

```{r}
#| label: fig-cluster_distribution
#| fig-cap: "Boxplots showing the distribution of indicator values in each cluster."
#| 
## Need a graph here (boxplot) showing the distribution of extracted SOC, pH, erosion, etc in each cluster.
```

```{r}
#| label: tbl-mclust-restoration
#| tbl-cap: "Classification of restoration effort based on Mclust classes."
#| 
kable(
  mclust_reclass |>
    mutate(restoration_effort = effort_labels[effort_class]),
  col.names = c("Mclust class", "Effort class", "Restoration effort")
)
```

```{r}
effort <- effort_raster()

if (!is.null(effort)) {
  area <- cellSize(effort, unit = "km")
  effort_area <- as.data.frame(c(effort, area), na.rm = TRUE)
  names(effort_area) <- c("effort_class", "area_km2")

  effort_summary <- effort_area |>
    group_by(effort_class) |>
    summarise(area_km2 = sum(area_km2), .groups = "drop") |>
    mutate(
      restoration_effort = effort_labels[effort_class],
      area_percent = area_km2 / sum(area_km2)
    ) |>
    select(restoration_effort, area_km2, area_percent)

  kable(effort_summary, digits = 2)
} else {
  cat("Attach or generate the Mclust cluster raster at: ", cluster_file)
}
```

```{r}
effort <- effort_raster()

if (!is.null(effort)) {
  plot(effort, col = effort_cols, main = "Reclassified Mclust Restoration Effort")
} else {
  cat("Attach or generate the Mclust cluster raster at: ", cluster_file)
}
```

## Random Forest Results

Three random forest models were used to evaluate low, medium, and high restoration effort. Each model produces a class-specific probability surface and a variable-importance graph showing which indicators contribute most to that class prediction.

### Low Restoration Effort

```{r}
if (file.exists(prob_files["Low"])) {
  plot(rast(prob_files["Low"]), main = "RF Probability: Low Restoration Effort")
} else {
  cat("Attach or generate the low-effort RF probability raster at: ", prob_files["Low"])
}
```

```{r}
if (file.exists(importance_pngs["Low"])) {
  include_graphics(importance_pngs["Low"])
} else {
  cat("Attach the low-effort variable-importance PNG at: ", importance_pngs["Low"])
}
```

### Medium Restoration Effort

```{r}
if (file.exists(prob_files["Medium"])) {
  plot(rast(prob_files["Medium"]), main = "RF Probability: Medium Restoration Effort")
} else {
  cat("Attach or generate the medium-effort RF probability raster at: ", prob_files["Medium"])
}
```

```{r}
if (file.exists(importance_pngs["Medium"])) {
  include_graphics(importance_pngs["Medium"])
} else {
  cat("Attach the medium-effort variable-importance PNG at: ", importance_pngs["Medium"])
}
```

### High Restoration Effort

```{r}
if (file.exists(prob_files["High"])) {
  plot(rast(prob_files["High"]), main = "RF Probability: High Restoration Effort")
} else {
  cat("Attach or generate the high-effort RF probability raster at: ", prob_files["High"])
}
```

```{r}
if (file.exists(importance_pngs["High"])) {
  include_graphics(importance_pngs["High"])
} else {
  cat("Attach the high-effort variable-importance PNG at: ", importance_pngs["High"])
}
```

## Country-Level Overview

At country level, the restoration-effort map shows the spatial distribution of low, medium, and high effort areas. The area graph summarizes how many square kilometers fall within each effort class.

```{r}
effort <- effort_raster()

if (!is.null(effort)) {
  plot(effort, col = effort_cols, main = "Somalia Restoration Effort")
} else {
  cat("Attach or generate the country restoration-effort map raster at: ", cluster_file)
}
```

```{r}
effort <- effort_raster()

if (!is.null(effort)) {
  area_df <- as.data.frame(c(effort, cellSize(effort, unit = "km")), na.rm = TRUE)
  names(area_df) <- c("effort_class", "area_km2")

  area_summary <- area_df |>
    group_by(effort_class) |>
    summarise(area_km2 = sum(area_km2), .groups = "drop") |>
    mutate(restoration_effort = factor(effort_labels[effort_class], effort_labels))

  ggplot(area_summary, aes(restoration_effort, area_km2, fill = restoration_effort)) +
    geom_col(width = 0.65) +
    scale_fill_manual(values = effort_cols) +
    labs(x = "Restoration effort class", y = "Area (km2)", fill = NULL) +
    theme_minimal() +
    theme(legend.position = "none")
} else {
  cat("Attach or generate the country restoration-effort raster at: ", cluster_file)
}
```

```{r}
if (exists("area_summary")) {
  kable(area_summary, digits = 2)
}
```

## Conclusion

The output shows that about 41% of Somalia is classified as medium restoration effort, 38% as low, and 21% as high. The variable importance graphs indicate that tree cover and pH are the most influential indicators for predicting restoration effort classes. These results can inform restoration planning and prioritization across Somalia.
